Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-04 paper/paper.Rmd
update codes and RJournal layout, not finishing...

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
blinded: 0
authors:

- name: H.Sherry Zhang
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of Econometrics and Business Statistics, Monash University 
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tour
- visualisation
- diagnostics 
- R
abstract: Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman & Tukey in 1974 stated that  "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
  \usepackage{todonotes}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      root.dir = here::here(), fig.path = here::here("figs/"), 
                      fig.align = 'center')
```

@@

Figure \@ref(fig:toy-search) shows a sample of the plot constructed using `search_better` with different values of the parameter `max_tries`. Comparing the returned index value of each iteration shows that the termination at `max_tries = 25` is not sufficient for `search_better` to explore the parameter space and a value of 500 is preferred over 25 in this context.

```{r toy-search, fig.cap = "(ref:toy-search)"}
```

(ref:toy-search) A comparison of `search_better` with different values of the parameter `max_tries`. A six-variable dataset `boa6` is used with `holes` index on a 2D problem. `max_tries = 500` allows the optimiser to find better basis with higher index value at iteration six.

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \@ref(fig:toy-interp) presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decrease pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost case is a deliberate construction of `search_better_random` where an inferior basis can be accepted with some probability so as to avoid being trapped in a local maximum.

@@

```{r noisy-better-geo,fig.asp=0.5,fig.cap = "(ref:noisy-better-geo)"}
```

@@

Figure \@ref(fig:noisy-better-geo) compares the tracing plot of two optimisers: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm and  `search_geodesic` fails to optimise this index with barely any improvement of the index value. On the other hand, `search_better` is doing relatively well. With the theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] producing an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(), raw_data = boa5) %>% tail(1) %>% pull(index_val)`, `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 3)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 3)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

@@

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is `boa6` where `x2` and `x7` are informative. The two theoretical best bases are [`r matrix(c(0, 1, 0, 0, 0, 0), nrow = 6)`] and [`r matrix(c(0, 0, 1, 0, 0, 0), nrow = 6)`] with index value `r index <- tourr::norm_kol(); proj <- as.matrix(boa6) %*% matrix(c(0, 1, 0, 0, 0, 0), nrow = 6); index(proj)` and `r index <- tourr::norm_kol(); proj <- as.matrix(boa6) %*% matrix(c(0, 0, 1, 0, 0, 0), nrow = 6); index(proj)`, respectively. Hence, the global maximum happens when variable `x7` is found. Simulation is done with 20 randomly generated seeds on two optimisers: `search_better` (Figure \@ref(fig:kol-better)) and `search_better_random` (Figure \@ref(fig:kol-random) and Figure \@ref(fig:kol-random-tuned)). Compared to Figure \@ref(fig:kol-better) and \@ref(fig:kol-random), Figure \@ref(fig:kol-random-tuned) further increases the neighbourhood parameter `alpha` from 0.5 to 0.7 to enlarge the search space.  The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

@@

>>>>

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
blinded: 0
author:
  - name: H.Sherry Zhang
    affiliation: Department of Econometrics and Business Statistics, Monash University 
    email:  Huize.Zhang@monash.edu
abstract: >
  Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman & Tukey in 1974 stated that  "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
keywords:
- optimisation
- projection pursuit 
- guided tour
- visualisation
- diagnostics 
- R
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
  \usepackage{todonotes}
output:
  bookdown::pdf_book:
    base_format: rticles::rjournal_article
    fig_height: 3
editor_options: 
  chunk_output_type: console

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      root.dir = here::here(), fig.path = here::here("figs/"), 
                      fig.align = 'center', cache = TRUE)
```

@@

Figure \ref{fig:toy-search} shows a sample of the plot constructed using `search_better` with different values of the parameter `max_tries`. Comparing the returned index value of each iteration shows that the termination at `max_tries = 25` is not sufficient for `search_better` to explore the parameter space and a value of 500 is preferred over 25 in this context.

```{r toy-search, fig.cap="\\label{fig:toy-search} A comparison of search\\_better with different values of the parameter max\\_tries. A six-variable dataset boa6 is used with holes index on a 2D problem. max\\_tries = 500 allows the optimiser to find better basis with higher index value at iteration six."}
```

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \ref{fig:toy-interp} presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decrease pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost case is a deliberate construction of `search_better_random` where an inferior basis can be accepted with some probability so as to avoid being trapped in a local maximum.

@@

<!-- ```{r noisy-better-geo,fig.asp=0.5,fig.cap = "(ref:noisy-better-geo)"} -->
<!-- ``` -->

@@

Figure \@ref(fig:noisy-better-geo) compares the tracing plot of two optimisers: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm and  `search_geodesic` fails to optimise this index with barely any improvement of the index value. On the other hand, `search_better` is doing relatively well. With the theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] producing an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(nrow(boa5)), raw_data = boa5) %>% tail(1) %>% pull(index_val)`, `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 3)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 3)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

@@

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is `boa6` where `x2` and `x7` are informative. The two theoretical best bases are [`r matrix(c(0, 1, 0, 0, 0, 0), nrow = 6)`] and [`r matrix(c(0, 0, 1, 0, 0, 0), nrow = 6)`] with index value `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 1, 0, 0, 0, 0), nrow = 6); index(proj)` and `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 0, 1, 0, 0, 0), nrow = 6); index(proj)`, respectively. Hence, the global maximum happens when variable `x7` is found. Simulation is done with 20 randomly generated seeds on two optimisers: `search_better` (Figure \@ref(fig:kol-better)) and `search_better_random` (Figure \@ref(fig:kol-random) and Figure \@ref(fig:kol-random-tuned)). Compared to Figure \@ref(fig:kol-better) and \@ref(fig:kol-random), Figure \@ref(fig:kol-random-tuned) further increases the neighbourhood parameter `alpha` from 0.5 to 0.7 to enlarge the search space.  The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

@@

\bibliography{RJreferences}
